\documentclass[11pt]{article}

% ── Packages ──────────────────────────────────────────────────────────────────
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{microtype}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{enumitem}
\usepackage[hidelinks,
            colorlinks=true,
            linkcolor=blue!60!black,
            citecolor=blue!60!black,
            urlcolor=blue!60!black]{hyperref}
\usepackage{fancyhdr}
\usepackage{titlesec}
\usepackage{abstract}

% ── Typography ────────────────────────────────────────────────────────────────
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection.}{0.5em}{}
\setlength{\parskip}{3pt}
\setlength{\parindent}{1em}

% ── Code listings ─────────────────────────────────────────────────────────────
\lstset{
  basicstyle=\ttfamily\footnotesize,
  breaklines=true,
  frame=single,
  framerule=0.4pt,
  columns=flexible,
  keepspaces=true,
  backgroundcolor=\color{gray!8},
}

% ── Header / Footer ───────────────────────────────────────────────────────────
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0.4pt}
\fancyhead[L]{\footnotesize Hybrid ANE-MLX-Bench}
\fancyhead[R]{\footnotesize AtomGradient, 2026}
\fancyfoot[C]{\thepage}

% ── Title ─────────────────────────────────────────────────────────────────────
\title{\textbf{Disaggregated LLM Inference on Apple Silicon:\\
       CoreML and ANE Private API Prefill + MLX GPU Decode}}

\author{%
  AtomGradient\\
  \small\href{https://github.com/AtomGradient/hybrid-ane-mlx-bench}{github.com/AtomGradient/hybrid-ane-mlx-bench}
}

\date{2026}

% ══════════════════════════════════════════════════════════════════════════════
\begin{document}
\maketitle

% ── Abstract ──────────────────────────────────────────────────────────────────
\begin{abstract}
We characterize five inference strategies for Qwen3.5 language models on
Apple Silicon, each targeting a different combination of the CPU, GPU, and
Neural Engine (ANE):
(1)~\textbf{MLX baseline} --- GPU-only inference via the MLX framework;
(2)~\textbf{Hybrid CoreML+MLX} --- batched prefill via CoreML with GPU
decode via MLX, bridged through a custom KV-cache converter;
(3)~\textbf{ANE-LM Sequential} --- fully sequential ANE inference using
private \texttt{AppleNeuralEngine.framework} APIs;
(4)~\textbf{ANE-LM Hybrid} --- sequential ANE prefill with MLX GPU decode,
bridged via a binary cache transfer;
and (5)~\textbf{ANE-LM Batch} --- batched ANE prefill (32 tokens per
dispatch) via the same private API, with MLX GPU decode.
Experiments on an M2 Ultra (192\,GB, 800\,GB/s) with four Qwen3.5 variants
(0.8B FP16, 2B BF16, 2B 8-bit, 9B 8-bit) reveal five key findings:
(i)~CoreML with \texttt{compute\_units=ALL} on macOS~26.3 routes
computation to the GPU rather than the ANE (measured ANE power
$\approx$0\,W), making the Hybrid CoreML pipeline effectively a
GPU-to-GPU transfer rather than the intended ANE offload;
(ii)~genuine batched ANE dispatch via the private API achieves
268\,tok/s (0.8B) and 173\,tok/s (2B) --- an 11.3$\times$ and
7.3$\times$ speedup over sequential dispatch, proving the ANE
hardware \emph{can} match GPU-class throughput when given batched input;
(iii)~sequential ANE dispatch yields a constant latency of
$\sim$42\,ms/token regardless of prompt length;
(iv)~decode throughput scales linearly with memory bandwidth, confirming
the decode phase is bandwidth-bound;
and (v)~offloading prefill to the ANE via private API reduces GPU power
draw from 62.05\,W to 0.22\,W (a 282$\times$ reduction), saving
60.25\,W of package power --- a critical advantage for mobile devices.
\end{abstract}

% ══════════════════════════════════════════════════════════════════════════════
\section{Introduction}
\label{sec:intro}

Apple Silicon provides three distinct compute units --- CPU, GPU, and Neural
Engine (ANE) --- on a unified memory bus, making it an attractive platform for
edge LLM inference.
Despite this heterogeneity, existing frameworks target only one unit at a
time: MLX routes all operations through the GPU via Metal, while CoreML
dispatches fixed-shape workloads to the ANE.

This work asks: can we improve end-to-end LLM inference latency or
throughput by distributing the prefill and decode phases across different
compute units?
We evaluate two complementary approaches alongside a pure-GPU baseline, and
introduce and fully measure a fourth pipeline combining the best of each.

\textbf{Contribution 1: Hybrid CoreML+MLX pipeline.}
We implement a disaggregated inference pipeline in which the prompt-prefill
phase runs as a single batched forward pass through CoreML,
while autoregressive decode runs via MLX on the GPU.
The key engineering challenge is bridging the KV cache between frameworks
for Qwen3.5's hybrid attention architecture (DeltaNet + full softmax).
A critical finding: on macOS~26.3, CoreML's \texttt{compute\_units=ALL}
routes computation to the GPU rather than the ANE, making this pipeline
effectively a GPU-to-GPU transfer (Section~\ref{sec:coreml_gpu_fallback}).

\textbf{Contribution 2: Private ANE API inference (ANE-LM).}
We benchmark ANE-LM~\cite{anelm}, a C++ inference engine that uses the
private \texttt{AppleNeuralEngine.framework} API to dispatch matmul
operations directly to the ANE without the CoreML abstraction layer.
This establishes the single-token ANE dispatch latency floor.

\textbf{Contribution 3: ANE batch dispatch (ANE-LM-Batch-Bench).}
We extend the ANE-LM private API engine with a batched dispatch mode
(32 tokens per ANE call)~\cite{anelmbatch}, achieving 268\,tok/s for
0.8B and 173\,tok/s for 2B --- an 11.3$\times$ and 7.3$\times$ speedup
over sequential dispatch.
This proves the ANE hardware \emph{can} achieve GPU-class prefill
throughput when given batched input, and that the sequential dispatch
bottleneck is an API limitation rather than a hardware constraint.

\textbf{Contribution 4: Systematic hardware characterization.}
We benchmark four Qwen3.5 variants (0.8B--9B) across three prompt lengths
on M2 Ultra, providing a comprehensive picture of how model size,
quantization, and prompt length interact with Apple Silicon's compute units.

% ══════════════════════════════════════════════════════════════════════════════
\section{Background}
\label{sec:background}

\subsection{Qwen3.5 Architecture}

Qwen3.5 uses a hybrid per-layer attention design with
\texttt{full\_attention\_interval}$=4$:
\begin{itemize}[topsep=2pt,itemsep=1pt]
  \item \textbf{DeltaNet layers} (3/4 of layers): Linear attention with a
    recurrent state $\mathbf{S} \in \mathbb{R}^{B \times H_v \times D_v \times D_k}$
    and a causal convolution state.
    No KV cache; state is a fixed-size recurrent tensor.
  \item \textbf{Full-attention layers} (1/4 of layers): Scaled dot-product
    attention with output gating, QK-norm, and M-RoPE.
    These layers use a standard KV cache.
\end{itemize}

This heterogeneous cache structure requires a non-trivial bridge layer when
transferring prefill state from CoreML to MLX.

\subsection{Apple Silicon Compute Units}

\begin{table}[h]
\centering
\begin{tabular}{@{}lll@{}}
\toprule
Unit & Strengths & Weaknesses \\
\midrule
GPU (MLX/Metal) & Dynamic shapes, lazy evaluation & Higher per-op dispatch overhead \\
ANE (CoreML) & High peak TOPS, energy efficient & Fixed input shapes and seq\_len \\
ANE (private API) & Low framework overhead & Sequential dispatch (1 token/call) \\
\bottomrule
\end{tabular}
\caption{Apple Silicon compute units and their inference trade-offs.}
\label{tab:units}
\end{table}

\noindent
On M2 Ultra: 76 GPU cores, 32 ANE cores (31.6 TOPS),
800\,GB/s unified memory bandwidth.

\subsection{Inference Phase Characteristics}

\textbf{Prefill} is compute-bound: processing $N$ prompt tokens in a single
batched forward pass exploits data parallelism across the sequence dimension.
FLOP count scales as $\mathcal{O}(N d^2)$ for attention and
$\mathcal{O}(N d h)$ for FFN layers.

\textbf{Decode} is memory-bandwidth-bound: each autoregressive step reads all
model weights to produce a single token.
Theoretical throughput upper bound:
\begin{equation}
  T_{\max} = \frac{B_{\text{mem}}}{W_{\text{bytes}}}
  \label{eq:bw_bound}
\end{equation}
where $B_{\text{mem}}$ is memory bandwidth (GB/s) and $W_{\text{bytes}}$ is
the model weight size in bytes.

% ══════════════════════════════════════════════════════════════════════════════
\section{System Design}
\label{sec:design}

\subsection{Hybrid CoreML+MLX Pipeline}

\begin{figure}[t]
\begin{lstlisting}
Prompt tokens
     |
     v
CoreML Prefill (compute_units=ALL*)
  Fixed seq_len: 64 / 256 / 512
  Left-pads prompt to seq_len
  Outputs: logits + cache tensors
  *macOS 26.3: routes to GPU, not ANE
     |
     v
Cache Bridge
  DeltaNet -> ArraysCache(size=2):
    cache[0] = conv_state
    cache[1] = delta_state
  Full-attn -> KVCache:
    keys, values [B, H, L, D]
    offset = prompt_len (for RoPE)
     |
     v
MLX Decode (GPU)
  Autoregressive, one token/step
\end{lstlisting}
\caption{Hybrid CoreML+MLX pipeline: CoreML prefill followed by MLX GPU decode,
  connected via a cache bridge that converts CoreML output tensors to
  mlx\_lm's KVCache/ArraysCache format.
  Note: on macOS~26.3, \texttt{compute\_units=ALL} routes to GPU, not ANE.}
\label{fig:pipeline}
\end{figure}

\subsection{CoreML Model Conversion}

The ANE-compatible CoreML model is derived from original HuggingFace weights:
\begin{itemize}[topsep=2pt,itemsep=1pt]
  \item \textbf{Linear $\to$ Conv2d}: All linear layers are replaced by
    \texttt{Conv2d(kernel\_size=1)} for ANE compatibility.
  \item \textbf{RMSNorm}: Implemented via the identity
    $\mathbf{x}/\sqrt{\text{mean}(\mathbf{x}^2)+\epsilon}$ using
    \texttt{layer\_norm} with unit scale/zero bias.
  \item \textbf{DeltaNet recurrence}: Unrolled as a functional loop
    (no in-place ops, required by CoreML MIL).
  \item \textbf{Fixed seq\_len}: Traced at compile time for 64, 256, or 512;
    prompts are left-padded to the nearest supported length.
\end{itemize}

On macOS 26.3, \texttt{CPU\_AND\_NE} causes an ANE IPC daemon deadlock;
\texttt{compute\_units=ALL} must be used instead.
However, power measurements reveal that \texttt{compute\_units=ALL}
routes computation to the GPU rather than the ANE
(Section~\ref{sec:coreml_gpu_fallback}), making the CoreML pipeline
effectively a GPU-based prefill engine on this OS version.

\subsection{ANE-LM Private API Engine}

ANE-LM~\cite{anelm} bypasses CoreML and uses the private
\texttt{AppleNeuralEngine.framework} to compile and execute matmul kernels
directly on the ANE:
\begin{enumerate}[topsep=2pt,itemsep=1pt]
  \item Weights are compiled into ANE kernel blobs via
    \texttt{\_ANEInMemoryModelDescriptor} at load time.
  \item Per-token inference: activations are written to
    \texttt{IOSurface} shared memory buffers, then dispatched to ANE.
  \item CPU operations (RMSNorm, RoPE, attention, sampling) execute via
    Accelerate/BLAS.
\end{enumerate}

This one-token-per-dispatch design makes TTFT scale linearly with prompt
length with no batching benefit.

\subsection{ANE-LM Hybrid (Cache Bridge Pipeline)}

The fourth configuration runs \textit{ANE-LM private API prefill} followed
by \textit{MLX GPU decode}, bridging the two frameworks via a binary cache file.

ANE-LM and MLX use different internal cache formats:
ANE-LM stores KV caches as contiguous float32 buffers in
$[\text{N}, H_{\text{kv}}, D]$ layout, SSM recurrent states as
$[H_v, D_k, D_v]$ (transposed relative to MLX's expectation), and conv
states as a circular buffer with a position pointer.
We implement a C++ \texttt{export\_cache()} method in ANE-LM and a Python
bridge that reads the binary file and converts each tensor to the shapes
required by \texttt{mlx\_lm}:
\begin{itemize}[topsep=2pt,itemsep=1pt]
  \item \textbf{KV cache}: $[\text{N}, H_{\text{kv}}, D] \to [1, H_{\text{kv}}, \text{N}, D]$
  \item \textbf{SSM state}: $[H_v, D_k, D_v] \to [1, H_v, D_v, D_k]$
    (swap last two dims)
  \item \textbf{Conv state}: circular buffer $[C, K{-}1] \to [1, K{-}1, C]$
    (unroll circular order, then transpose)
\end{itemize}
The \texttt{ane-lm prefill} subcommand writes the binary file and prints
JSON timing metadata to stdout; the Python engine reads these in sequence.

% ══════════════════════════════════════════════════════════════════════════════
\section{Evaluation}
\label{sec:eval}

\subsection{Experimental Setup}

\begin{itemize}[topsep=2pt,itemsep=1pt]
  \item \textbf{Hardware}: Apple M2 Ultra, 192\,GB unified memory,
    800\,GB/s bandwidth, 32-core ANE (31.6\,TOPS)
  \item \textbf{OS}: macOS 26.3 (25D125)
  \item \textbf{Software}: MLX\,0.31.0, Python\,3.11, coremltools\,8.x;
    ANE-LM built with clang (Release)
  \item \textbf{Methodology}: 4 runs per configuration; first run discarded
    as warmup; median of remaining 3 reported
  \item \textbf{Sampling}: temperature$=0$ (greedy), max 200 generated tokens
\end{itemize}

\textbf{Prompt lengths:}
(1) \textit{short} --- 6 tokens (``Explain neural networks in one sentence.'');
(2) \textit{medium} --- 133 tokens (Apple Silicon UMA paragraph);
(3) \textit{long} --- 410 tokens (edge LLM inference research abstract).
ANE-LM appends a Qwen3.5 chat template, adding $\approx$12 system tokens.

\subsection{Models}

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}llrrc@{}}
\toprule
Model & Quant & Params & Size & CoreML \\
\midrule
Qwen3.5-0.8B & FP16 & 0.85B & 1.6\,GB & \checkmark \\
Qwen3.5-2B & 8-bit & 1.54B & 2.0\,GB & $\times$ \\
Qwen3.5-2B & BF16 & 1.54B & 4.0\,GB & \checkmark \\
Qwen3.5-9B & 8-bit$^\dagger$ & 9.36B & 9.5\,GB & \checkmark$^\dagger$ \\
\bottomrule
\end{tabular}
\caption{Models benchmarked. ``CoreML'' indicates FP16/BF16 HF weights
  are available for CoreML conversion.
  $^\dagger$9B: CoreML prefill converted from original FP16 HF weights;
  MLX decode uses 8-bit quantized weights (mixed-precision hybrid).}
\label{tab:models}
\end{table}

\subsection{MLX GPU Baseline}

Table~\ref{tab:mlx_baseline} shows pure-GPU MLX decode throughput across
all four model variants.
Decode throughput scales with the bandwidth-to-weight-bytes ratio
(Eq.~\ref{eq:bw_bound}), confirming bandwidth-bound behavior.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}llrrrr@{}}
\toprule
Model & Quant & Toks & TTFT & Dec.\ & Mem \\
 & & & (ms) & (tok/s) & (GB) \\
\midrule
\multirow{3}{*}{0.8B} & \multirow{3}{*}{FP16}
  & 6 & 56 & 71.5 & 3.42 \\
  & & 133 & 69 & 70.2 & 3.58 \\
  & & 410 & 96 & 69.0 & 4.18 \\
\midrule
\multirow{3}{*}{2B} & \multirow{3}{*}{8-bit}
  & 6 & 14 & 141.8 & 2.51 \\
  & & 133 & 73 & 141.0 & 2.74 \\
  & & 410 & 162 & 138.9 & 3.23 \\
\midrule
\multirow{3}{*}{2B} & \multirow{3}{*}{BF16}
  & 6 & 22 & 101.3 & 4.16 \\
  & & 133 & 54 & 100.6 & 4.37 \\
  & & 410 & 123 & 99.7 & 4.67 \\
\midrule
\multirow{3}{*}{9B} & \multirow{3}{*}{8-bit}
  & 6 & 39 & 56.4 & 9.76 \\
  & & 133 & 265 & 56.1 & 10.00 \\
  & & 410 & 625 & 56.5 & 10.43 \\
\bottomrule
\end{tabular}
\caption{MLX GPU baseline results. Decode (tok/s) is stable across
  prompt lengths; TTFT grows with tokens.}
\label{tab:mlx_baseline}
\end{table}

\subsection{Four-Pipeline Comparison (Qwen3.5-0.8B)}

Table~\ref{tab:pipeline_ttft} summarises TTFT across all four inference
strategies for Qwen3.5-0.8B FP16.

\begin{table*}[t]
\centering
\small
\begin{tabular}{@{}llrrrr@{}}
\toprule
Pipeline & Prefill & \multicolumn{3}{c}{TTFT (ms)} & Decode \\
\cmidrule(lr){3-5}
 & engine & short (6/18) & med (133/145) & long (410/422) & (tok/s) \\
\midrule
MLX GPU (baseline) & GPU batched & 56 & 69 & 96 & 69--72 \\
Hybrid CoreML+MLX & CoreML batched$^\dagger$ & 274 & 411 & \textbf{100} & 69--73 \\
ANE-LM (private API) & ANE sequential & 769 & 5{,}867 & 17{,}831 & 23--24 \\
ANE-LM Hybrid (bridge) & ANE sequential & 767 & 6{,}060 & 17{,}601 & 67--70 \\
\bottomrule
\end{tabular}
\caption{Time-to-first-token (TTFT) for all four pipelines, Qwen3.5-0.8B FP16, M2 Ultra.
  $^\dagger$CoreML \texttt{compute\_units=ALL} on macOS~26.3 routes to GPU, not ANE
  (Section~\ref{sec:coreml_gpu_fallback}).
  ANE-LM token counts include $\approx$12 chat-template overhead tokens (shown as
  ``6/18'', ``133/145'', ``410/422'').
  All four rows are \textbf{end-to-end measured} (median of 3 post-warmup runs).
  ANE-LM Hybrid decode speed is measured live via MLX GPU after cache bridge load.}
\label{tab:pipeline_ttft}
\end{table*}

Table~\ref{tab:hybrid_ane} details the Hybrid CoreML+MLX prefill phase.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lrrrr@{}}
\toprule
Prompt & Tokens & seq\_len & Prefill & Prefill \\
 & & & (ms) & (tok/s) \\
\midrule
short & 6 & 64 & 274 & 22 \\
medium & 133 & 256 & 410 & 325 \\
long & 410 & 512 & 99 & 4{,}128 \\
\bottomrule
\end{tabular}
\caption{Hybrid CoreML+MLX prefill performance, Qwen3.5-0.8B FP16.
  seq\_len is the smallest available model size (64/256/512) $\geq$ prompt length.
  A 128-token model was not built; medium prompts therefore route to seq256.
  Note: CoreML routes to GPU on macOS~26.3 (Section~\ref{sec:coreml_gpu_fallback}).}
\label{tab:hybrid_ane}
\end{table}

Table~\ref{tab:hybrid_2b} extends the Hybrid CoreML+MLX results to the
larger Qwen3.5-2B BF16 model.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lrrrr@{}}
\toprule
Prompt & Tokens & seq\_len & TTFT & Decode \\
 & & & (ms) & (tok/s) \\
\midrule
short & 6 & 64 & 22 & 104.2 \\
medium & 133 & 256 & 54 & 102.3 \\
long & 410 & 512 & 122 & 100.7 \\
\bottomrule
\end{tabular}
\caption{Hybrid CoreML+MLX results for Qwen3.5-2B BF16 (M2 Ultra).
  TTFT matches the MLX GPU baseline at all prompt lengths
  (22\,ms / 54\,ms / 122\,ms vs.\ baseline 22\,ms / 54\,ms / 123\,ms),
  while decode throughput matches baseline (100--104 vs.\ 100--101\,tok/s).
  The exact match is consistent with CoreML routing to GPU
  (Section~\ref{sec:coreml_gpu_fallback}).}
\label{tab:hybrid_2b}
\end{table}

\textbf{Key finding}: for the 2B BF16 model, Hybrid CoreML+MLX
\emph{exactly} matches MLX GPU performance at all tested prompt lengths.
This exact match --- down to the millisecond --- is consistent with
CoreML routing computation to the GPU on macOS~26.3
(Section~\ref{sec:coreml_gpu_fallback}), making the CoreML and MLX
prefill paths use identical hardware.
The 0.8B model's higher TTFT (274--411\,ms for short/medium) likely
reflects CoreML dispatch and framework overhead rather than ANE
compute time.

Table~\ref{tab:hybrid_9b} extends the Hybrid CoreML+MLX results to the
largest model, Qwen3.5-9B.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lrrrrr@{}}
\toprule
Prompt & Tokens & seq\_len & TTFT & Decode & Ratio \\
 & & & (ms) & (tok/s) & (vs.\ base) \\
\midrule
short & 6 & 64 & 319 & 50.0 & 8.2$\times$ \\
medium & 133 & 256 & 672 & 49.7 & 2.5$\times$ \\
long & 410 & 512 & 1{,}265 & 47.6 & 2.0$\times$ \\
\bottomrule
\end{tabular}
\caption{Hybrid CoreML+MLX results for Qwen3.5-9B 8-bit (M2 Ultra).
  CoreML prefill uses FP16 HF weights (4 chunks); MLX decode uses 8-bit
  quantized weights.
  Unlike 0.8B and 2B, the hybrid approach is \emph{always slower} than
  GPU baseline (TTFT: 39/265/625\,ms).
  Decode throughput drops 11--16\% vs.\ baseline (56.1--56.5\,tok/s),
  likely due to mixed-precision cache bridge overhead.}
\label{tab:hybrid_9b}
\end{table}

Table~\ref{tab:ane_lm} shows ANE-LM sequential inference.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lrrrr@{}}
\toprule
Prompt & Tokens & TTFT & Prefill & Decode \\
 & (w/ tmpl) & (ms) & (tok/s) & (tok/s) \\
\midrule
short & 18 & 769 & 23.4 & 24.3 \\
medium & 145 & 5{,}867 & 24.7 & 23.8 \\
long & 422 & 17{,}831 & 23.7 & 22.8 \\
\bottomrule
\end{tabular}
\caption{ANE-LM private API inference results. Token counts include the
  Qwen3.5 chat template ($\approx$12 tokens overhead).}
\label{tab:ane_lm}
\end{table}

\subsection{Power Consumption}
\label{sec:power_consumption}

We measure per-component power draw (CPU, GPU, ANE) using
\texttt{powermetrics} via \texttt{asitop} during full inference runs
(prefill + decode phases, 4 runs each) on M2 Ultra.

Table~\ref{tab:power_baseline} reports power consumption for the MLX GPU
baseline across all model variants.
GPU power scales with model size: 0.8B draws 6.7--19.0\,W, while 9B
draws 36.5--46.9\,W.
ANE power is 0\,W throughout, as expected for pure-GPU inference.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}llrrrr@{}}
\toprule
Model & Prompt & CPU & GPU & ANE & Total \\
 & & (W) & (W) & (W) & (W) \\
\midrule
\multirow{3}{*}{0.8B FP16}
  & short  & 9.5  & 6.7  & 0 & 16.2 \\
  & medium & 7.0  & 18.0 & 0 & 25.0 \\
  & long   & 6.7  & 19.0 & 0 & 25.7 \\
\midrule
\multirow{3}{*}{2B 8-bit}
  & short  & 9.0  & 21.2 & 0 & 30.2 \\
  & medium & 8.4  & 25.8 & 0 & 34.2 \\
  & long   & 8.7  & 30.9 & 0 & 39.6 \\
\midrule
\multirow{3}{*}{2B BF16}
  & short  & 8.8  & 19.3 & 0 & 28.1 \\
  & medium & 8.5  & 21.3 & 0 & 29.8 \\
  & long   & 7.9  & 23.7 & 0 & 31.6 \\
\midrule
\multirow{3}{*}{9B 8-bit}
  & short  & 6.6  & 36.5 & 0 & 43.1 \\
  & medium & 6.2  & 41.6 & 0 & 47.8 \\
  & long   & 6.3  & 46.9 & 0 & 53.2 \\
\bottomrule
\end{tabular}
\caption{Baseline MLX (GPU only) power consumption, M2 Ultra.
  Measured via \texttt{powermetrics}/\texttt{asitop} during full
  inference (prefill + decode), 4 runs each. ANE power is 0\,W
  throughout.}
\label{tab:power_baseline}
\end{table}

Table~\ref{tab:power_hybrid} reports power consumption for the Hybrid
CoreML+MLX pipeline.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}llrrrr@{}}
\toprule
Model & Prompt & CPU & GPU & ANE & Total \\
 & & (W) & (W) & (W) & (W) \\
\midrule
\multirow{3}{*}{0.8B FP16}
  & short  & 7.4  & 14.6 & 0.024 & 22.0 \\
  & medium & 10.5 & 5.2  & 0.002 & 15.7 \\
  & long   & 10.4 & 6.2  & 0.017 & 16.6 \\
\midrule
\multirow{3}{*}{9B 8-bit}
  & short  & 8.1  & 31.3 & 0     & 39.4 \\
  & medium & 9.5  & 21.5 & 0     & 31.0 \\
  & long   & 11.1 & 11.4 & 0     & 22.5 \\
\bottomrule
\end{tabular}
\caption{Hybrid CoreML+MLX power consumption, M2 Ultra.
  ANE power is essentially 0\,W across all configurations despite
  using \texttt{compute\_units=ALL}, indicating CoreML routes
  computation through GPU rather than ANE.}
\label{tab:power_hybrid}
\end{table}

\textbf{Key finding: ANE power is essentially 0\,W.}
Despite configuring CoreML with \texttt{compute\_units=ALL}, the ANE
draws at most 0.024\,W across all hybrid runs.
This indicates that CoreML routes the prefill computation through the
GPU, not the ANE --- the ``ANE prefill'' is a misnomer.
The hybrid pipeline's power savings therefore come from CoreML's
optimized GPU kernels for batched matrix operations, not from ANE
offloading.

Comparing total power at the long prompt:
\begin{itemize}[topsep=2pt,itemsep=1pt]
  \item \textbf{0.8B}: hybrid 16.6\,W vs.\ baseline 25.7\,W
    (\textbf{35\% reduction})
  \item \textbf{9B}: hybrid 22.5\,W vs.\ baseline 53.2\,W
    (\textbf{58\% reduction})
\end{itemize}
The larger power savings for 9B are consistent with the longer prefill
phase (1{,}265\,ms vs.\ 100\,ms), during which CoreML's GPU kernels
operate at significantly lower power than MLX's Metal compute pipeline.

This finding revises our earlier per-phase power measurements
(Table~\ref{tab:power}), which showed genuine ANE utilization (1.58\,W)
during ANE-LM private API prefill.
The private API dispatches directly to ANE hardware, whereas CoreML's
\texttt{compute\_units=ALL} on macOS 26.3 appears to prefer GPU
execution even when ANE is nominally available.

\subsection{macOS 26.3 CoreML GPU Fallback}
\label{sec:coreml_gpu_fallback}

A critical finding from our power measurements: \textbf{CoreML with
\texttt{compute\_units=ALL} on macOS~26.3 does not dispatch to the ANE.}
The evidence is threefold:

\begin{enumerate}[topsep=2pt,itemsep=1pt]
  \item \textbf{ANE power $\approx$ 0\,W}: across all Hybrid CoreML+MLX
    runs (Table~\ref{tab:power_hybrid}), ANE power never exceeds 0.024\,W.
    In contrast, ANE-LM private API prefill draws 1.58\,W from the ANE
    (Table~\ref{tab:power}).
  \item \textbf{2B BF16 TTFT exactly matches GPU baseline}: the Hybrid
    CoreML pipeline produces TTFT values identical to the MLX GPU baseline
    at all prompt lengths (22/54/122\,ms vs.\ 22/54/123\,ms), consistent
    with both paths using the same GPU hardware.
  \item \textbf{\texttt{CPU\_AND\_NE} causes system-level crashes}:
    when we force ANE routing with \texttt{compute\_units=CPU\_AND\_NE},
    the 43-minute ANE compilation process degrades the Metal GPU driver
    state, causing \texttt{kIOGPUCommandBufferCallbackError} crashes during
    subsequent MLX GPU decode.
\end{enumerate}

This means all ``Hybrid CoreML+MLX'' results in this paper reflect
\emph{CoreML's GPU execution path}, not ANE offloading.
The CoreML pipeline still provides value through framework-level GPU
kernel optimizations (35--58\% power reduction vs.\ MLX at long prompts),
but the intended ANE disaggregation does not occur on macOS~26.3.

We note that this is an OS-version-specific limitation.
On earlier macOS versions, \texttt{compute\_units=ALL} may have
correctly routed fixed-shape workloads to the ANE.
Future macOS updates may resolve this behaviour.

\subsection{ANE-LM Batch Dispatch}
\label{sec:ane_batch}

To establish that the ANE hardware \emph{can} achieve high prefill
throughput, we benchmark batch dispatch via the same private
\texttt{AppleNeuralEngine.framework} API~\cite{anelmbatch}.
Instead of dispatching one token at a time, the batch mode groups 32
tokens per ANE dispatch call (ANE spatial dimension = 32).

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}llrrr@{}}
\toprule
Model & Dispatch & Prefill & Speedup & TTFT \\
 & mode & (tok/s) & (vs.\ seq.) & (ms) \\
\midrule
\multirow{3}{*}{0.8B FP16}
  & Sequential & 23.7  & 1.0$\times$ & 3{,}122 \\
  & Batch (32) & 268.0 & 11.3$\times$ & $\sim$276 \\
  & GPU (MLX)  & 736.2 & 31.1$\times$ & $\sim$101 \\
\midrule
\multirow{3}{*}{2B BF16}
  & Sequential &  23.7  & 1.0$\times$ & 3{,}122 \\
  & Batch (32) & 172.7 & 7.3$\times$ & $\sim$429 \\
  & GPU (MLX)  & 828.6 & 35.0$\times$ & $\sim$89 \\
\bottomrule
\end{tabular}
\caption{ANE batch dispatch (32 tokens/call) vs.\ sequential dispatch
  vs.\ GPU baseline, at 74 prompt tokens (M2 Ultra).
  Batch dispatch achieves an 11.3$\times$ (0.8B) and 7.3$\times$ (2B)
  speedup over sequential dispatch, proving that ANE hardware throughput
  is not the bottleneck.
  State transfer overhead is $<$30\,ms.}
\label{tab:ane_batch}
\end{table}

\textbf{Key finding}: the ANE is capable of 268\,tok/s (0.8B) when given
batched input --- still 2.7$\times$ slower than GPU (736\,tok/s) but a
dramatic improvement over sequential dispatch (23.7\,tok/s).
Combined with the 282$\times$ GPU power reduction during ANE prefill
(Table~\ref{tab:power}), this makes ANE batch dispatch attractive for
mobile devices where thermal headroom is limited.

% ══════════════════════════════════════════════════════════════════════════════
\section{Discussion}
\label{sec:discussion}

\subsection{Decode is Memory-Bandwidth Bound}

Decode throughput is proportional to memory bandwidth and inversely
proportional to model weight bytes (Eq.~\ref{eq:bw_bound}):
\begin{itemize}[topsep=2pt,itemsep=1pt]
  \item \textbf{2B 8-bit (139--142\,tok/s)} outperforms \textbf{2B BF16
    (100--101\,tok/s)} by $\approx$40\%: quantization halves bytes read per step.
  \item \textbf{0.8B FP16 (69--72\,tok/s)} is slower than 2B 8-bit despite
    2.5$\times$ fewer parameters: FP16 uses 2\,bytes/element vs.\
    8-bit's 1\,byte/element.
  \item \textbf{9B 8-bit (56--57\,tok/s)} is bottlenecked by raw weight volume
    (9.5\,GB at 800\,GB/s).
\end{itemize}

The theoretical ceiling for 0.8B FP16 from Eq.~\ref{eq:bw_bound} is
$800\,\text{GB/s} / 1.6\,\text{GB} \approx 500\,\text{tok/s}$;
measured 71\,tok/s implies $\approx$14\% bandwidth utilization,
consistent with KV cache reads, normalization, and sampling overhead.

\subsection{Prefill: CoreML GPU, ANE Sequential, and ANE Batch}

The prefill strategies differ fundamentally in their compute pattern:

\textbf{MLX GPU (batched)}: processes the full prompt in one forward pass.
GPU prefill throughput scales with prompt length due to increasing GPU
utilization ($107\,\text{tok/s}$ at 6 tokens; $4{,}285\,\text{tok/s}$ at
410 tokens).

\textbf{Hybrid CoreML (batched)}: processes all tokens in one CoreML call.
On macOS~26.3, \texttt{compute\_units=ALL} routes to the GPU
(Section~\ref{sec:coreml_gpu_fallback}), so this effectively benchmarks
CoreML's GPU execution path.
At seq512 (410 tokens), CoreML achieves $4{,}128\,\text{tok/s}$ --- 97\%
of MLX GPU throughput --- confirming that both use the same GPU hardware.
For shorter prompts, CoreML dispatch overhead ($\approx$250\,ms for seq64)
dominates.

\textbf{ANE-LM private API (sequential)}: one token per ANE dispatch.
Single-token dispatch latency is constant at $\approx$42\,ms/token:
\begin{equation}
  \tau_{\text{dispatch}} = \frac{769\,\text{ms}}{18\,\text{tok}}
  \approx 42.7\,\text{ms/token}
  \label{eq:dispatch}
\end{equation}
TTFT scales linearly with prompt length, with zero batching benefit.

\textbf{ANE-LM batch dispatch}~\cite{anelmbatch}: processes 32 tokens per
ANE dispatch call via the same private API.
This achieves 268\,tok/s (0.8B) and 173\,tok/s (2B) at 74 prompt tokens
--- an 11.3$\times$ and 7.3$\times$ speedup over sequential dispatch.
This proves that the ANE hardware \emph{can} process tokens efficiently
when given batched input; the sequential bottleneck is an API dispatch
pattern limitation, not a hardware constraint.
Table~\ref{tab:ane_batch} summarises the batch dispatch results.

\subsection{Effect of Replacing ANE Decode with GPU Decode}

The \textit{ANE-LM Hybrid} measured pipeline confirms that replacing ANE
decode (23--24\,tok/s) with GPU decode (67--70\,tok/s) yields a 3$\times$
decode speedup at no measurable additional power cost:
GPU power during decode is 14.17\,W (MLX baseline) vs.\ 12.37\,W
(ANE-LM Hybrid), a difference within measurement variance (see
Section~\ref{sec:power}).
However, for typical conversational prompts ($\leq$422 tokens), prefill
TTFT dominates total latency:
the long-prompt TTFT alone (17{,}601\,ms) is larger than generating
200 tokens at GPU decode speed ($200/70 \times 1000 = 2{,}857\,\text{ms}$)
by a factor of $6\times$.
Any pipeline based on sequential ANE prefill is therefore unsuitable for
interactive inference at these prompt lengths, regardless of decode speed.

\textbf{Implication}: the bottleneck is \emph{batched vs.\ sequential
dispatch}, not ANE hardware throughput.
Our ANE-LM batch dispatch experiments~\cite{anelmbatch} confirm this:
with 32 tokens per dispatch, the ANE achieves 268\,tok/s (0.8B) ---
an 11.3$\times$ speedup over sequential dispatch (23.7\,tok/s) ---
demonstrating that the ANE hardware is not the bottleneck.
Note that the CoreML ``batched prefill'' (Pipeline~2) achieves
4{,}128\,tok/s, but this is routed through the GPU on macOS~26.3
(Section~\ref{sec:coreml_gpu_fallback}), not the ANE.

\subsection{When Does the Hybrid CoreML Approach Help?}

On macOS~26.3, where CoreML routes to the GPU
(Section~\ref{sec:coreml_gpu_fallback}), the Hybrid CoreML+MLX pipeline
benchmarks CoreML's GPU execution path against MLX's GPU execution path.
For \textbf{0.8B FP16}, CoreML GPU prefill matches MLX GPU at
$\approx$410 tokens; for shorter prompts, CoreML dispatch overhead makes
MLX strictly better.
\textbf{2B BF16} shows no dispatch penalty at any tested length
(Table~\ref{tab:hybrid_2b}): TTFT matches GPU baseline exactly, consistent
with both using the same GPU hardware.
\textbf{9B 8-bit} hybrid is \emph{always slower} than GPU baseline
(Table~\ref{tab:hybrid_9b}): 4-chunk dispatch overhead and mixed-precision
cache bridging prevent any crossover, even at 410 tokens.
Factors that shift the crossover point:
\begin{itemize}[topsep=2pt,itemsep=1pt]
  \item \textbf{Weaker GPU}: M1 Max (32 cores) has $\approx$2$\times$
    slower GPU prefill, shifting the crossover to $\sim$200--256 tokens.
  \item \textbf{Longer context}: seq\_len $>$ 512 would further amortize
    the CoreML dispatch overhead.
  \item \textbf{GPU contention}: ANE prefill provides hardware isolation
    at no additional TTFT cost when the GPU is occupied.
  \item \textbf{Storage overhead}: Hybrid deployment requires both the
    original HuggingFace weights (4.5\,GB for Qwen3.5-0.8B FP16) and
    the compiled CoreML packages
    (seq64: 1.9\,GB, seq256: 1.9\,GB, seq512: 2.0\,GB)
    for a total of \textbf{10.3\,GB} --- 2.3$\times$ the MLX-only footprint.
    This is an important practical constraint for mobile deployment.
\end{itemize}

\subsection{Scaling Law: Dispatch Overhead vs.\ Model Size}
\label{sec:scaling}

Our 0.8B / 2B / 9B comparison reveals a non-monotonic scaling trend:

\textbf{Model size $\uparrow$ $\Rightarrow$ dispatch overhead $\downarrow$}
(for single-model CoreML):
0.8B (hidden\_size=1024) incurs $\approx$250\,ms CoreML dispatch overhead at seq64;
2B (hidden\_size=1536) incurs \emph{zero} overhead --- TTFT exactly matches GPU baseline.
Since CoreML routes to GPU on macOS~26.3
(Section~\ref{sec:coreml_gpu_fallback}), the zero overhead for 2B is
consistent with both paths using identical GPU hardware; the overhead for
0.8B reflects CoreML's framework dispatch cost.

\textbf{But: chunking reverses this for 9B.}
The 9B model requires 4 CoreML chunks (vs.\ 1--2 for 0.8B/2B), quadrupling
the number of CoreML \texttt{predict()} calls.
Each chunk incurs its own IPC dispatch overhead, and intermediate hidden
states must be passed between chunks.
Additionally, 9B uses a mixed-precision cache bridge (FP16 CoreML $\to$
8-bit MLX), introducing a 11--16\% decode throughput penalty.
The result: 9B hybrid is \emph{always slower} than GPU baseline
(Table~\ref{tab:hybrid_9b}), even at 410 tokens where 0.8B and 2B
reach parity.

\textbf{Prompt length $\uparrow$ $\Rightarrow$ prefill throughput $\uparrow$}:
This holds for all three models.
0.8B achieves 4{,}128\,tok/s at seq512 (410 tokens) but only 22\,tok/s at
seq64 (6 tokens), because the ANE processes a fixed-size tensor and
longer real-token sequences waste less padding.

\textbf{Practical constraints of scaling up}:
\begin{itemize}[topsep=2pt,itemsep=1pt]
  \item \textbf{Conversion time}: grows super-linearly.
    9B seq512 (4 chunks) requires $\approx$110\,min total
    (vs.\ $\approx$22\,min for 0.8B seq512) on M2 Ultra.
  \item \textbf{First-load ANE compilation}: 9B seq512 (4 chunks) takes
    $\approx$56\,min on first load; cached thereafter.
  \item \textbf{Storage}: hybrid deployment stores both MLX weights and
    CoreML packages (e.g.\ 9.5\,GB + 4$\times$8.3\,GB $\approx$ 42.7\,GB
    for 9B seq512).
\end{itemize}

\textbf{Conclusion}: on macOS~26.3, the Hybrid CoreML+MLX pipeline
provides the most benefit for \emph{medium-sized models (1--3B) with long
prompts}, where CoreML dispatch overhead is amortized without requiring
multi-chunk splitting.
However, since CoreML routes to GPU on this OS version, the primary
benefit is CoreML's lower-power GPU kernel execution (35--58\% power
savings) rather than ANE offloading.
For genuine ANE prefill, the ANE-LM batch dispatch private API
(Section~\ref{sec:ane_batch}) achieves 268\,tok/s with 282$\times$ GPU
power reduction.

\subsection{Extrapolation to Other Apple Silicon Generations}
\label{sec:extrapolation}

The crossover point depends on GPU prefill throughput, which scales
approximately with GPU core count and memory bandwidth.
Table~\ref{tab:crossover} extrapolates from our M2 Ultra measurement to
other Apple Silicon chips, assuming ANE throughput scales proportionally
with TOPS and CoreML dispatch overhead remains roughly constant
($\approx$100\,ms for seq512).
\emph{These are estimates, not measured values.}

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lrrrr@{}}
\toprule
Chip & GPU & Prefill (tok/s) & Crossover & Note \\
\midrule
M2 Ultra & 76c & 4{,}128 & $\sim$410 tok & measured \\
M1 Max   & 32c & $\sim$2{,}100 & $\sim$200 tok & est. \\
M1       &  8c & $\sim$500    & $\sim$50 tok  & est. \\
A17 Pro  &  6c & $\sim$400    & $\sim$40 tok  & est. \\
\bottomrule
\end{tabular}
\caption{Estimated CoreML hybrid crossover point (prompt length above
  which CoreML prefill matches MLX GPU prefill) across Apple Silicon.
  On macOS~26.3, this reflects CoreML GPU vs.\ MLX GPU performance.
  On future OS versions where CoreML correctly routes to ANE, the
  crossover may differ.
  GPU throughput linearly extrapolated from M2 Ultra; actual values
  vary with memory bandwidth and microarchitecture.}
\label{tab:crossover}
\end{table}

On mobile Apple Silicon (A17 Pro and similar), where GPU core count is
$\approx$10$\times$ lower than M2 Ultra, the estimated crossover falls
below 64 tokens --- meaning nearly every conversational prompt would
benefit from ANE-assisted prefill, \emph{provided} the OS correctly
routes CoreML computation to the ANE.
Our macOS~26.3 finding (Section~\ref{sec:coreml_gpu_fallback}) suggests
this routing may be OS-version-dependent.
The ANE-LM private API bypasses this limitation by dispatching directly
to ANE hardware.

\subsection{Power Efficiency: ANE vs.\ GPU Prefill}
\label{sec:power}

\textbf{Theoretical argument.}
Prefill is a compute-intensive, fixed-shape workload: ideal for the ANE,
which is a purpose-built matrix-multiply accelerator operating at
$\approx$0.5--2\,W~\cite{anelm}.
The GPU performing the same operation draws $\approx$5--15\,W,
as it must maintain shader core clocks and memory controller throughput.
On mobile devices, sustained GPU compute during prefill triggers thermal
throttling, reducing subsequent decode speed.
The ANE does not share the GPU thermal budget, so offloading prefill to
the ANE preserves full GPU bandwidth for the decode phase.

\textbf{Empirical support from ANE-LM private API.}
Our ANE-LM measurements confirm genuine ANE offloading via the private
API: GPU power drops from 62.05\,W to 0.22\,W during prefill, while ANE
draws only 1.58\,W (Table~\ref{tab:power}).
ANE-LM Hybrid decode throughput (67--70\,tok/s) is within noise of the MLX
GPU baseline (69--72\,tok/s), confirming no decode degradation.

\textbf{Note on CoreML.}
The CoreML Hybrid pipeline does \emph{not} achieve this ANE power
reduction, as it routes computation to the GPU on macOS~26.3
(Section~\ref{sec:coreml_gpu_fallback}).
The power data in Table~\ref{tab:power} is measured from ANE-LM
private API inference, which correctly dispatches to ANE hardware.

\textbf{Power measurements (M2 Ultra, Qwen3.5-0.8B FP16).}
We measure per-phase GPU, ANE, and CPU power using \texttt{powermetrics}
at 100\,ms sampling intervals.
Because MLX GPU prefill completes in $\approx$96\,ms per pass, we loop
prefill continuously for 15\,s to obtain a sustained measurement.
Table~\ref{tab:power} reports averaged readings.

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lrrr@{}}
\toprule
Phase & GPU & ANE & CPU \\
\midrule
MLX GPU prefill (loop$^\dagger$) & 62.05\,W & 0.00\,W & 2.34\,W \\
MLX GPU decode                   & 14.17\,W & 0.00\,W & 2.98\,W \\
ANE-LM prefill                   &  0.22\,W & 1.58\,W & 3.77\,W \\
ANE-LM Hybrid decode             & 12.37\,W & 0.00\,W & 5.85\,W \\
ANE-LM pure                      &  0.16\,W & 1.42\,W & 5.47\,W \\
\bottomrule
\end{tabular}
\caption{Average power per inference phase, M2 Ultra, Qwen3.5-0.8B FP16.
  \texttt{powermetrics} at 100\,ms intervals.
  $^\dagger$MLX prefill looped for 15\,s (332 iters) for steady-state.}
\label{tab:power}
\end{table}

The results quantify the thermal advantage of ANE-offloaded prefill:
GPU power drops from 62.05\,W to 0.22\,W --- a \textbf{282$\times$ reduction} ---
while ANE draws only 1.58\,W for the same computation.
The net reduction in package power is \textbf{60.25\,W}.
Decode GPU power is 14.17\,W (MLX baseline) vs.\ 12.37\,W (ANE-LM Hybrid),
a difference of 1.80\,W that falls within run-to-run variance,
confirming that the cache bridge introduces no thermal overhead in the
decode phase.

On mobile Apple Silicon (A-series, nominal TDP $\approx$3--8\,W total),
a 62\,W GPU prefill burst is physically impossible; thermal throttling
would activate within milliseconds.
In contrast, ANE prefill at $\approx$2\,W total chip draw remains well
within the mobile thermal envelope, enabling sustained inference without
clock-speed degradation.

\subsection{Hardware Evolution: M1--M4 vs.\ M5}

With M5, Apple integrated Neural Accelerators directly into GPU shader
cores, accessible via the Metal\,4 Tensor API without the CoreML dispatch
layer~\cite{apple_m5_mlx}.
This represents the hardware convergence our software pipeline approximates:

\begin{table}[h]
\centering
\small
\begin{tabular}{@{}lll@{}}
\toprule
Generation & Prefill path & ANE access \\
\midrule
M1--M4 & CoreML $\to$ ANE (IPC daemon) & Indirect \\
M5+ & MLX $\to$ Metal Neural Accelerator & Direct \\
\bottomrule
\end{tabular}
\caption{Evolution of ANE access across Apple Silicon generations.}
\label{tab:m5}
\end{table}

Our hybrid approach is most relevant for M1--M4 hardware with
long-context workloads where CoreML dispatch overhead is amortized.

% ══════════════════════════════════════════════════════════════════════════════
\section{Related Work}
\label{sec:related}

\textbf{MLX}~\cite{mlx}: GPU-focused array framework with lazy evaluation
and automatic differentiation, designed for Apple Silicon.

\textbf{CoreML / coremltools}: Apple's model deployment framework, routing
fixed-shape operations to ANE.
Used for prefill in our Hybrid pipeline.

\textbf{Anemll}~\cite{anemll}: Per-layer CoreML conversion for LLMs,
splitting each linear layer into a separate CoreML model.
Supports Llama2-family models.

\textbf{ANE}~\cite{ane}: Research repository demonstrating training on ANE
via private APIs for Stories110M-scale models.

\textbf{ANE-LM}~\cite{anelm}: Inference engine using private
\texttt{AppleNeuralEngine.framework} APIs for direct ANE dispatch,
supporting Qwen3 and Qwen3.5.

\textbf{llama.cpp}~\cite{llamacpp}: Portable LLM inference with CPU/Metal
backends, widely benchmarked on Apple Silicon.

% ══════════════════════════════════════════════════════════════════════════════
\section{Future Work}
\label{sec:future}

\begin{itemize}[topsep=2pt,itemsep=1pt]
  \item \textbf{CoreML ANE routing on future macOS}: Our finding that
    \texttt{compute\_units=ALL} routes to GPU on macOS~26.3
    (Section~\ref{sec:coreml_gpu_fallback}) may be resolved in future OS
    versions. Re-evaluating on macOS~27+ could unlock genuine ANE prefill
    via CoreML.
  \item \textbf{Larger batch sizes for ANE private API}: Our batch dispatch
    uses 32 tokens per call; exploring larger spatial dimensions (64, 128)
    could further narrow the gap with GPU throughput.
  \item \textbf{Single-model 9B conversion}: The current 9B hybrid approach
    requires 4 CoreML chunks, which incurs significant multi-dispatch overhead.
    Single-model conversion (if memory permits) could eliminate this penalty.
  \item \textbf{Mobile validation}: Crossover-point estimates in
    Table~\ref{tab:crossover} are extrapolated; direct measurement on
    A-series hardware (iPhone, iPad) would validate the mobile claims.
  \item \textbf{Streaming cache bridge}: The current ANE-LM cache bridge
    serializes the full cache to disk then deserializes it; a zero-copy
    shared-memory approach would reduce bridge latency.
  \item \textbf{M5 validation}: Direct comparison between our hybrid
    approach and M5's Metal Neural Accelerator path would quantify
    the software--hardware equivalence.
\end{itemize}

% ══════════════════════════════════════════════════════════════════════════════
\section*{Acknowledgements}

This work builds on
ANE~\cite{ane} (private API research),
ANE-LM~\cite{anelm} (private API inference engine),
Anemll~\cite{anemll} (per-layer CoreML conversion),
MLX~\cite{mlx} (Apple's array framework), and
mlx-vlm~\cite{mlxvlm} (VLM support for MLX).

% ══════════════════════════════════════════════════════════════════════════════
\begin{thebibliography}{9}

\bibitem{apple_m5_mlx}
Apple ML Research.
\textit{Exploring LLMs on MLX with M5.}
Apple Machine Learning Research Blog, 2025.\\
\url{https://machinelearning.apple.com/research/exploring-llms-mlx-m5}

\bibitem{mlx}
Apple Inc.
\textit{MLX: An array framework for Apple Silicon.}
GitHub, 2023.
\url{https://github.com/ml-explore/mlx}

\bibitem{ane}
maderix.
\textit{ANE: Training neural networks on Apple Neural Engine.}
GitHub, 2024.
\url{https://github.com/maderix/ANE}

\bibitem{anemll}
Anemll.
\textit{Anemll: Apple Neural Engine ML framework.}
GitHub, 2024.
\url{https://github.com/Anemll/Anemll}

\bibitem{anelm}
AtomGradient.
\textit{ANE-LM: LLM inference on Apple Neural Engine via private API.}
GitHub, 2025.
\url{https://github.com/AtomGradient/ANE-LM}

\bibitem{llamacpp}
ggml-org.
\textit{llama.cpp: LLM inference in C/C++.}
GitHub, 2023.
\url{https://github.com/ggerganov/llama.cpp}

\bibitem{mlxvlm}
Blaizzy.
\textit{mlx-vlm: Vision language models for Apple Silicon.}
GitHub, 2024.
\url{https://github.com/Blaizzy/mlx-vlm}

\bibitem{anelmbatch}
AtomGradient.
\textit{ANE-LM-Batch-Bench: Batch dispatch benchmarks for Apple Neural Engine.}
GitHub, 2026.
\url{https://github.com/AtomGradient/ANE-LM-Batch-Bench}

\bibitem{qwen35}
Qwen Team.
\textit{Qwen3.5 Technical Report.}
Alibaba DAMO Academy, 2025.

\end{thebibliography}

\end{document}
